Hydras \& Co.

### 2022-02-15 doc/gaia-chapter.tex
initialization

\section{Introduction}
The \gaia project by Jos\'e Grimm aimed to formalize mathematics in \coq in
the style of Nicolas Bourbaki. The formalization of the first book in the
Elements of Mathematics series by Bourbaki, on the theory of sets, was
initially described in a technical report in July 2009~\cite{Grimm2009a}.
The set-theoretic axioms and basic definitions in \gaia were derived
from an earlier development by Carlos Simpson~\cite{Simpson2004,CatsZFCContrib}.
Grimm then wrote (and continually updated) technical reports describing the
formalization of Bourbaki's two subsequent books~\cite{Grimm2009b,Grimm2016}
and additional topics in number theory~\cite{grimm:hal-00911710,Grimm2014},
before he passed away in 2019.

In 2020, members of \community transferred the \gaia source code to
GitHub and adapted it for recent releases of the Mathematical Components
library, which \gaia heavily relies on.
Anonymous volunteers (``collaborators of Nicolas Bourbaki'') then finished
the only in-progress proof left by Grimm. At around 155,000 LOC, \gaia is currently one of the largest maintained open source \coq projects~\cite{Gaia}.

@@

\begin{itemize}
\item From \gaia, we import the proof of associativity of multiplication of ordinals strictly less than $\epsilon_0$, which was missing from \Hydras.
  \item The other way, we started to import definitions and properties of canonical sequences and rapidly growing functions from \Hydras (see Chapter~\ref{chap:ketonen} and\ref{chap:alpha-large}~of this book).
\end{itemize}

Please note that by ``import'' we do not mean ``duplicate the proofs''.  Whenever possible, we proceed by rewriting steps over the statements and not the proofs.

This initial draft bridge code mostly uses the SSReflect proof language
and idioms from the Mathematical Components library. We made this design
decision since we believe it is less challenging to reason about
Hydra-battles code using SSReflect and MathComp than to reason about
Gaia without SSReflect.

\section{Library structure}
The \gaiaHydras library is designed for use  by users of
\gaia and/or  \HydrasLib. Whenever possible, we managed to respect the notations and naming conventions of these libraries.

In case of name clash, we often added a one-letter prefix to names : \texttt{g} or \texttt{G} for \gaia, \texttt{h} or \texttt{H}
  for \HydrasLib.

For instance, both \gaia and \HydrasLib have a type called \texttt{T1} (data type for Cantor normal forms).
\begin{itemize}
\item In
  \href{../theories/html/gaia_hydras.T1Bridge_use.html}%
  {\texttt{gaia\_hydras.T1Bridge\_use}}, we adopt \HydrasLib's point of view, and the name \texttt{T1} is by default bound to
\HydrasLib' data structures.

@@

Please note that in this module, we favour \gaia's notations over hydra's. A snippet like \texttt{alpha < succ beta} should be interpreted as \texttt{T1lt alpha beta}, where \texttt{T1lt} is defined in ~\href{https://github.com/coq-community/gaia/blob/master/theories/ssete9.v}{\texttt{gaia.ssete9.v}}.

@@

\subsection{Examples of refinements}
Both libraries define constants like $0$, $1$, $\omega$, and arithmetic functions: successor, addition, multiplication, and exponential of base $\omega$ (function $\phi_0$). We prove consistency of these definitions. Finally, we prove that the boolean predicates `` to be in normal form'' are equivalent.

@@

\subsection{Importing a theorem from \gaia}
The \texttt{Gaia} library already contains a proof that the multiplication of ordinals less than $\epsilon_0$ is associative.
\emph{From~\href{https://github.com/coq-community/gaia/blob/master/theories/ssete9.v}{gaia.ssete9.v}}

@@

\subsection{Importing theorems from \HydrasLib}

@@

\subsubsection{Canonical sequences}

In~\href{../theories/html/gaia_hydras.GCanon.html}%
{\texttt{gaia\_hydras.GCanon}}, canonical sequences are just imported from the library \linebreak
\href{../theories/html/hydras.Epsilon0.Canon.html}%
{\texttt{hydras.Epsilon0.Canon}}.

First, we rename \HydrasLib' \texttt{canon} function into
\texttt{hcanon}. Then we define a new function \emph{via} the
translations \texttt{g2h} and \texttt{h2g}.

@@

The following lemmas are proved by rewriting from corresponding statements proved in \HydrasLib.
\emph{The author of these proof scripts is still a beginner in Ssreflect. Please forgive the clumsiness of the scripts below.}

@@

@@

\section{Well formed ordinals}
For sake of compatibility, we add a clone of \texttt{hydra-battles}' type \texttt{E0} defined in \href{../theories/html/hydras.Epsilon0.E0.html}{Epsilon0.E0}.

>>>>

\section{Introduction}
The \gaia project~\cite{Gaia} by Jos\'e Grimm aimed to formalize mathematics in \coq  in
the style of Nicolas Bourbaki. The formalization of the first book in the
Elements of Mathematics series by Bourbaki, on the theory of sets, was
initially described in a technical report in July 2009~\cite{Grimm2009a}.
The set-theoretic axioms and basic definitions in \gaia were derived
from an earlier development by Carlos Simpson~\cite{Simpson2004,CatsZFCContrib}.
Grimm then wrote (and continually updated) technical reports describing the
formalization of Bourbaki's two subsequent books~\cite{Grimm2009b,Grimm2016}
and additional topics in number theory~\cite{grimm:hal-00911710,Grimm2014},
before he passed away in 2019.

In 2020, members of \community transferred the \gaia source code to
GitHub and adapted it for recent releases of the Mathematical Components
library, which \gaia heavily relies on.
Anonymous volunteers (``collaborators of Nicolas Bourbaki'') then finished
the only in-progress proof left by Grimm. At around 155,000 LOC, \gaia is currently one of the largest maintained open source \coq projects.

@@

\begin{itemize}
\item \gaia contains lemmas on ordinal arithmetic (\emph{e.g.} associativity of multiplication of ordinals strictly less than $\epsilon_0$), which were missing from \Hydras. 
  \item The other way, \Hydras contains formal definitions and proofs of properties of canonical sequences and rapidly growing functions  (see Chapters~\ref{chap:ketonen} and~\ref{chap:alpha-large}~of this book).
  \end{itemize}

  We plan to make accessible the first kind of lemmas to \HydrasLib' users and the second kind to \gaia's. For this purpose, we write modules dedicated to the importation of definitions and lemmas from each of both libraries into the other one.
Please note that by ``import'' we do not mean ``duplicate the proofs''.  Whenever possible, we proceed by rewriting steps over the statements and not the proofs.

This initial draft bridge code mostly uses the SSReflect proof language
and idioms from the Mathematical Components library. We made this design
decision since we believe it is less challenging to reason about
Hydra-battles code using SSReflect and MathComp than to reason about
Gaia without SSReflect \footnote{The author of these proof scripts is still a beginner in Ssreflect. Please forgive the clumsiness of the current proof scripts.}.

\section{Library structure}
The \gaiaHydras library is designed for use  by users of
\gaia and/or  \HydrasLib. Whenever possible, we managed to respect the notations and naming conventions of both libraries.

In case of name clash (for instance the type \texttt{T1} of ordinal terms below $\epsilon_0$), we often added a one-letter prefix to names : \texttt{g} or \texttt{G} for \gaia, \texttt{h} or \texttt{H}
for \HydrasLib.

In the following two examples, the bare name \texttt{T1} is given by default either  to \HydrasLib' world or \gaia's.

\begin{itemize}
\item In
  \href{../theories/html/gaia_hydras.T1Bridge_use.html}%
  {\texttt{gaia\_hydras.T1Bridge\_use}}, we adopt \HydrasLib's point of view, and the name \texttt{T1} is by default bound to
\HydrasLib' data structures.

@@

Please note that in this module, we favour \gaia's notations over hydra's. A snippet like \texttt{(alpha < succ beta)} should be interpreted as \texttt{(T1lt alpha beta)}, where \texttt{T1lt} is defined in ~\href{https://github.com/coq-community/gaia/blob/master/theories/ssete9.v}{\texttt{gaia.ssete9.v}}.

@@

\subsection{Examples of refinements}
Both of our libraries define constants like $0$, $1$, $\omega$, and arithmetic functions: successor, addition, multiplication, and exponential of base $\omega$ (function $\phi_0$). We prove consistency of these definitions. Finally, we prove that the boolean predicates `` to be in normal form'' are equivalent.

@@

\section{Importing a theorem from \gaia}
The \texttt{Gaia} library already contains a proof that the multiplication of ordinals less than $\epsilon_0$ is associative.
\emph{From~\href{https://github.com/coq-community/gaia/blob/master/theories/ssete9.v}{gaia.ssete9.v}}

@@

\section{Importing theorems from \HydrasLib}

@@

\subsection{Canonical sequences}

In~\href{../theories/html/gaia_hydras.GCanon.html}%
{\texttt{gaia\_hydras.GCanon}}, canonical sequences are just imported from the library \linebreak
\href{../theories/html/hydras.Epsilon0.Canon.html}%
{\texttt{hydras.Epsilon0.Canon}}.
First, we rename \HydrasLib' \texttt{canon} function into
\texttt{hcanon}. Then we define a new function \emph{via} the
translations \texttt{g2h} and \texttt{h2g}.

@@

The following lemmas are proved by rewriting from corresponding statements proved in \HydrasLib.

@@

\subsection{A family of rapidly growing functions}

The functions $F_\alpha$ are described in Section~\vref{sect:wainer}. The symbol \texttt{F\_} has type \texttt{E0 -> nat -> nat},
where \texttt{E0} is the type of well formed ordinal terms less than $\epsilon_0$.
\subsubsection{Well-formed ordinal terms}

For sake of compatibility, we add a clone of \texttt{hydra-battles}' type \texttt{E0} defined in \href{../theories/html/hydras.Epsilon0.E0.html}{Epsilon0.E0}.

\subsubsection{Importing the \texttt{F\_alpha} family}

We are now able to import \HydrasLib' $F_\alpha$ functions.

\inputsnippets{GF_alpha/FAlphaDef}

Mainly by rewriting, we import a few properties of the $F_\alpha$s.

\inputsnippets{T1Bridge/MonoDef, GF_alpha/FAlphaProps}.

@@
